---
title: "Temporal variation in transmission during the COVID-19 outbreak in Italy"
csl: resources/bmj.csl
output:
  pdf_document: default
  html_document:
    theme: cosmo
bibliography: resources/library.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE,
                      fig.width = 6, fig.height = 3,
                      message = FALSE,
                      warning = FALSE,
                      dpi = 320,
                      fig.path = "figures/")
```


```{r load-packages, include = FALSE}
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(lubridate)
library(patchwork)
library(ggplot2)
library(knitr)
library(kableExtra)
```

```{r settings}
## Report date
target_date <- Sys.Date() - 1

## Countries to include - based on folder names
regions <- list.files("../results/")

## Put into alphabetical order
regions <- regions[order(regions)]

names(regions) <- regions 

names(regions)[regions %in% "national"] <- "National"

## Results directory (relative to report directory)
results_dir <- "../results"

## News updates
updates <- list.files("../man/news") %>% 
  stringr::str_remove(".md") %>% 
  as.Date(.) %>% 
  sort()

## Generic function to load data
load_data <- function(file = NULL, country = NULL, date = target_date,
                      result_dir = results_dir) {
  file_path <- file.path(result_dir, country, date, file)
  object <- readRDS(file_path)
  
  return(object)
}

## Overall figures and tables
summary_figures <- 3
summary_tables <- 1
regional_methods_dir <- "man/region-specific-methods"
news_dir <- "man/news"
```

*S. Abbott (1),  J. D. Munday (1), J. Hellewell (1), R. N. Thompson (1), N. Bosse (1), CMMID COVID team (1), S. Flasche (1),  A. J. Kucharski (1), R. M. Eggo (1), S. Funk (1).*

*Correspondence to: sam.abbott@lshtm.ac.uk* 

*1. Center for the Mathematical Modelling of Infectious Diseases, London School of Hygiene & Tropical Medicine, London WC1E 7HT, United Kingdom*

*Last Updated:* `r target_date`

*Note: this is preliminary analysis, has not yet been peer-reviewed and is updated daily as new data becomes available. This work is licensed under a [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/). A summary of this report can be downloaded [here](https://cmmid.github.io/topics/covid19/current-patterns-transmission/reports/national-time-varying-transmission/italy.pdf)*

## Summary {.tabset}

**Aim:** To identify changes in the reproduction number, rate of spread, and doubling time during the course of the COVID-19 outbreak in Italy whilst accounting for potential biases due to delays in case reporting.


`r paste0("**Latest estimates as of the ", target_date, "**")`

### Region map

```{r load-results}
map_prob_change <- function(var) {
      dplyr::case_when(var < 0.05 ~ "Increasing",
                       var < 0.2 ~  "Likely increasing", 
                       var < 0.8 ~ "Unsure", 
                       var < 0.95 ~ "Likely decreasing",
                       var <= 1 ~ "Decreasing") %>% 
        factor(levels = c("Increasing", "Likely increasing", "Unsure", 
                          "Likely decreasing", "Decreasing"))
}

estimates <- tibble::tibble(
  `Country/Region` = names(regions),
  `Cases with date of onset on the day of report generation` = regions %>% 
     purrr::map_chr(~ load_data("current_cases.rds", .)),
  `Expected change in daily cases` = regions %>% 
  purrr::map_dbl(~ load_data("prob_control_latest.rds", .)) %>% 
  map_prob_change(),
  `Effective reproduction no.` =  regions %>% 
  purrr::map_chr(~ load_data("bigr_eff_latest.rds", .)),
  `Doubling time (days)` = regions %>% 
  purrr::map_chr(~ load_data("doubling_time_latest.rds", .))) 
```


```{r extract-numeric-results}
## Extract a value
extract_var <- function(var, index) {
  var %>% 
      str_split(" -- ") %>% 
      purrr::map_dbl(~ as.numeric(.[[index]]))
}

## Make estimates numeric
numeric_estimates <- estimates %>% 
  dplyr::select(`Country/Region`, `Cases with date of onset on the day of report generation`, 
                `Effective reproduction no.`, `Expected change in daily cases`) %>% 
  tidyr::gather(value = "value", key = "metric", -`Country/Region`, -`Expected change in daily cases`) %>% 
  dplyr::mutate(
    lower = extract_var(value, 1),
    upper = extract_var(value, 2))

numeric_estimates <- numeric_estimates %>% 
  dplyr::mutate(
        `Country/Region` = numeric_estimates$`Country/Region` %>% 
      factor(levels = numeric_estimates %>% 
               arrange(desc(upper)) %>% 
               pull(`Country/Region`) %>% 
               unique())
  )

## Highest inc countires
high_inc_regions <- numeric_estimates %>% 
  dplyr::arrange(dplyr::desc(upper)) %>% 
  pull(`Country/Region`) %>% 
  unique() %>% 
  as.character() 

regions <- purrr::map_chr(high_inc_regions, ~ regions[names(regions) %in% .])
names(regions) <- high_inc_regions
```

```{r national-map, fig.width = 6, fig.height = 6}
## Load case data with region codes
region_codes <- readRDS("../data/region_codes.rds")

estimates %>% 
  dplyr::rename(region = `Country/Region`) %>%  
  dplyr::left_join(region_codes, by = "region") %>% 
  EpiNow::country_map(country = "Italy", variable = "Expected change in daily cases") +
  ggplot2::guides(fill = ggplot2::guide_legend(title = "Expected change in daily cases",
                                               ncol = 2))
```

<br>
`r paste0("*Figure 1: Regional map of the expected change in daily cases based on data from the ", target_date, ".*")`

### Summary of latest reproduction number and case count estimates

```{r, fig.height = 12, fig.width = 12}
numeric_estimates %>% 
  ggplot2::ggplot(ggplot2::aes(x = `Country/Region`, ymin = lower, 
                               ymax = upper, col = `Expected change in daily cases`)) +
  ggplot2::geom_linerange(size = 6) +
  ggplot2::geom_hline(yintercept = 1, linetype = 2) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  cowplot::theme_cowplot() +
  cowplot::panel_border() +
  theme(axis.text.x = element_text(angle = 90)) +
  ggplot2::scale_color_viridis_d(option = "cividis", end = 0.9,alpha = 0.7) +
  theme(legend.position = "bottom") +
  ggplot2::guides(col = ggplot2::guide_legend(nrow = 2))
```

<br>
`r paste0("*Figure 2: Cases with date of onset on the day of report generation and the time-varying estimate of the effective reproduction number (bar = 95% credible interval) based on data from the ", target_date, ". Regions are ordered by the number of expected daily cases and shaded based on the expected change in daily cases. The dotted line indicates the target value of 1 for the effective reproduction no. required for control and a single case required fror elimination.*")`

### Reproduction numbers over time in the 5 regions with the most cases currently and nationally

```{r summarise-bigr-plot, fig.height = 9, fig.width = 12}
plots <- regions[names(regions) %in% high_inc_regions[1:6]] %>% 
  purrr::map(function(region) {
    plot <- load_data("bigr_eff_plot.rds", region) +
      ggplot2::labs(title = region %>% 
             stringr::str_replace("-", " ") %>% 
    stringr::str_to_title()) +
    ggplot2::scale_x_date(date_breaks = "1 week", date_labels = "%b %d",
                          limits = c(as.Date("2020-03-01"), as.Date(target_date))) +
    ggplot2::coord_cartesian(ylim = c(0, 4))
    
    return(plot)
  })

plots %>% 
  patchwork::wrap_plots() +
  patchwork::plot_layout(ncol = 2)
```

<br>
`r paste0("*Figure 3: Time-varying estimate of the effective reproduction number (light grey ribbon = 95% credible interval; dark grey ribbon = the interquartile range) based on data from the ", target_date, " in the regions expected to have the highest number of incident cases. Confidence in the estimated values is indicated by shading with reduced shading corresponding to reduced confidence. The dotted line indicates the target value of 1 for the effective reproduction no. required for control.*")`


### Latest estimates summary table

```{r summary-table-1}
caption <- paste0("Latest estimates of the number of cases by date of onset, the effective reproduction number, and the doubling time for the ", target_date, " in each region included in the analysis. Based on the last 7 days of data. The 95\\% credible interval is shown for each numeric estimate.")

estimates <- names(regions) %>% 
  purrr::map_dfr(~ dplyr::filter(estimates, `Country/Region` %in% .))

knitr::kable(estimates, booktabs = TRUE, caption = ifelse(!knitr:::is_latex_output(), "", caption))  %>% 
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"),
                full_width = TRUE) %>% 
  kableExtra::landscape()
```
<br>
`r if(!knitr:::is_latex_output()) paste0("*Table 1: ", caption, "*")`


## Methods {.tabset}

### Summary

* Case counts by date, stratified by region, were constructed from daily datasets made publically available by the Dipartimento della Protezione Civile [@italydata, @NCoVUtils].
* Case onset dates were estimated using case counts by date of report and a distribution of reporting delays fitted to a European line-list.
* Censoring of cases was adjusted for by assuming that the number of cases is drawn from a binomial distribution.
* Time-varying effective reproduction estimates were made with a 7-day sliding window using *EpiEstim* [@EpiEstim; @THOMPSON2019100356] adjusted for imported cases and assuming an uncertain serial interval with a mean of 4.7 days (95% CrI: 3.7, 6.0) and a standard deviation of 2.9 days (95% CrI: 1.9, 4.9) [@Nishiura2020.02.03.20019497].
* Time-varying estimates of the doubling time were made with a 7-day sliding window by iteratively fitting an exponential regression model. 
* The methods in this report are based on our previous study of the global temporal variation during the COVID-19 outbreak [@timevary].

### Limitations

* The estimated onset dates are based on current European data for the delay in reporting. These data may not be representative of the underlying reporting distribution.
* The estimate of not-yet-confirmed cases to scale up recent numbers is uncertain and relies on the observed delays to confirmation to remain constant over the course of the outbreak.
* All data used is at a national/regional level; diagnostic capabilities may vary in different parts of each region, adding uncertainty to the reported numbers.
* Trends identified using our approach are robust to under-reporting assuming it is constant but absolute values may be biased by reporting rates. Pronouced changes in reporting rates may also impact the trends identified.
* Data on imported cases was not available (either international imports or between region imports).

### Detail

#### Data

We used a European line-list that contained the date of symptom onset, date of confirmation and import status (imported or local) for each case [@kraemer2020epidemiological, @NCoVUtils] where available. Daily case counts by date of report and region were extracted from daily datasets made publically available by the Dipartimento della Protezione Civile [@italydata, @NCoVUtils]. 


#### Statistical analysis

We used the same approach as in our previous global study of the temporal variation in transmission during the COVID-19 outbreak [@timevary]. However, due to a limited line-list of Italian cases we used a combined linelist of cases from Germany, France, Italy, Austria, the Netherlands, Belgium, and Spain to estimate the report delay. We could also not account for imported cases (either international or between region) due to a shortage of data. 


`r if(!knitr:::is_latex_output()) '## Regional reports {.tabset}'`

```{r generate-region-reports, include = FALSE, eval = !knitr:::is_latex_output()}
region_summaries <- 1:length(regions) %>% 
  purrr::map(function(region_index) {
    index <- region_index
    region <- regions[region_index]
    region_name <- names(regions)[region_index]
   out <- knitr::knit_child(system.file("rmarkdown-templates/region-report.Rmd", 
                                        package = "EpiNow"), envir = environment()) 
  })
```


```{r print-region-reports, results="asis", eval = !knitr:::is_latex_output()}
cat(paste(region_summaries, collapse = '\n'))
```

`r if(!knitr:::is_latex_output()) '## Updates {.tabset}'`

```{r generate-update-reports, include = FALSE, eval = !knitr:::is_latex_output()}
update_reports <- rev(updates) %>% 
  purrr::map(function(target_update) {
    update <- target_update
   out <- knitr::knit_child(system.file("rmarkdown-templates/report-news.Rmd",
                                        package = "EpiNow"), envir = environment()) 
  })
```

```{r print-updates, results="asis", eval = !knitr:::is_latex_output()}
cat(paste(update_reports, collapse = '\n'))
```

## References

<div id = 'refs'></div>
